One-step, three-factor authentication with in-ear EEG

### 2016-12-06 paper.tex
Minor (final?) fixes

We propose the use of custom-fit Ear EEG technology as the platform for
investigating the feasibility, performance, and usability of one-step three-factor
authentication. In addition to the same knowledge factor and inherence factor as
in our previous work, the user can submit in the same step the possession factor
in the form of the EEG-sensing ear-piece(s) that are custom-fitted to and worn in
their ear. These earpieces can serve as physical tokens in the same way as bank
ATM cards and wearable hardware tokens. Furthermore, because the earpieces
are custom-fitted to each individual, they will likely not be able to produce good
electrical impedances when worn by a different individual.

@@

We propose an update and expansion of the authentication tasks, to take
advantage of our lessons learned from previous studies on the relative strengths
of different mental tasks, with regards to authentication accuracy and usability as
reported by participants. Furthermore, given the in-ear placement of the
electrodes, and hence spatial proximity to the temporal lobes, we see a unique
opportunity to design and test novel authentication tasks based on either auditory
imagery or auditory steady-state response (ASSR). The ten authentication tasks and their
attributes are listed in the tables below.

@@

This will allow us to quantify the security and usability performances of these
tasks, both collectively as they contribute to overall authentication accuracy, as
well as comparatively against one another.
We will use OpenBCI, a open-source EEG biosensing system as our data
collection platform. This requires integrating the custom-fit Ear EEG devices with
the OpenBCI board. We will also need to extend our own data collector
software, to support the recording of OpenBCI data that is synchronized with
PsychoPy, which we use for stimuli presentation. We will recruit on the order of
20-30 participants who are willing to commit to multiple session visits, first for ear
molding, then for device impedance testing, and then for one or more data
collection sessions.

@@

Our initial participants were recruited from a nearby university and scheduled for ear molding
and impedance checking sessions. Finally, the data collection visit was scheduled and took
approximately 90 minutes for set up and experiment execution. The OpenBCI system we used
allows for 8 channels of simultaneous recording, along with separate ground and reference channels.
Data was initially collected with the ground placed at the center of the forehead, and using the left
mastoid as reference, though we can easily re-reference to another channel by subtracting a desired
channel (such as right mastoid). Each earpiece (shown in the image below) contain three channels each: 
one placed on the helix, and two inside the canal - one front-facing and the other back-facing. The remaining
two channels were placed on the right mastoid for later re-referencing, and at approximately Fp1 (on the 
forehead above the left eye) for validating the data collected in the ears against a scalp-based measure. 
Before beginning the experiment, the data from all channels was visualized and participants were asked to
blink and clench their jaws to confirm all channels were active and properly connected.

During the experiment, participants were seated in a comfortable position in a quiet room facing a laptop screen on which the
instructions and stimuli were presented. Each task was completed once in sets five trials each, and then
each was completed again for another five trials. Each trial was 10 seconds in length, for a total of 10 trials and 
100 seconds of data collected per task. The instructions were read aloud to the participant by the experimenter, and
the experiment was advanced using a pointer held in the participant's lap to minimize motion artifacts in the data.  
The experimenter also recorded the participant's chosen secrets for the sport, song, face, speech, and sequence
 tasks and reminded the participant of these for the second set of trials.

@@

The recorded impedances of the earpiece electrodes were less than 5 kOhms, a benchmark used widely in previous
ear EEG work, except one. The left helix electrode of one participant was measured at 9 kOhms, and generally the 
helix impedances for both participants were higher than their ear canal counterparts. We expected this result, given that the
helix electrode relies on quality of the the earpiece's fit outside the ear, and is not as securely and tightly placed as the
electrodes within the ear canal. Nonetheless, the data from all electrodes were tested in the remaining two data quality tests.

For the alpha-attenuation test, data from the "Breathe" task was compared with that of the "Breathe - Open" task. It is a well-
known feature of EEG data that activity in the alpha-band (approximately 8-12 Hz range) increases when the eyes are closed
compared with a similar state with eyes open. For both of our pilot participants this attenuation is clearly visible even in just
a single trial's data. To further validate, we also performed this calculation on the data collected from the Fp1 electrode and see
the effect clearly here as well. It is important to note that the left ear results are reported using the right mastoid as reference, and
the right ear results in turn using the left mastoid as reference. When using the same side mastoid for reference the effect is not 
visible, though may be if we average across many trials. This is not surprising, as the further a reference electrode is from the active
channel the less "real" signal is being subtracted from the active channel. This has important design implications for eventual 
real-world deployment of this authentication method however, as it will likely involve pieces on or around both ears to properly 
function, and not just one. The figures below show the alpha attenuation in the left and right ear channels, as well as Fp1.

@@

\begin{figure}[h]
  \centering
  \begin{minipage}[b]{0.45\textwidth}
    \includegraphics[width=\textwidth]{001_ASSR_Left.jpg}
  \end{minipage}
  \hfill
  \begin{minipage}[b]{0.45\textwidth}
    \includegraphics[width=\textwidth]{001_ASSR_Right.jpg}
  \end{minipage}
\caption{Power spectra for data collected from the earpiece channels during 40 Hz ASSR stimulus. Again, the spike is clearly visible though not at 40 Hz, however it 
does match the }
\end{figure}

@@

Following [5], we use logarithmic binning to produce
compressed feature vectors of a variable size. This technique
has been show to offer robust, linear classifiability in healthy
subjects. It is unique in its use of the entire frequency
spectrum. Since EEG activity is associated with frequencies
from 1-40Hz, we presume this range contains the majority of
relevant signal. However, we do not rule out the possibility
that useful signal exists in other frequency ranges. Muscular
activity, for example, might be correlated with mental gestures
in some cases. Logarithmic binning produces feature vectors
biased toward known sources of signal, while still including
data points from outside this frequency range that may be
informative.

@@

Though our classifier's accuracy is strong, we still cannot say that it relies 
on both knowledge and inherence to authenticate the subject. Is it the passthought
that the classifier is authenticating, or do users (or earphones) simply have characteristic
EEG readings?

@@

In expanding on the feasibility study we describe above, 
our first task comprises of collecting data from more subjects,
ideally for a subject population of 30.
In addition, we hope to collect more diverse data,
by taking regular samples over a period of a few weeks,
and in a variety of circumstances. Our goal here is gague the stability
of EEG readings over time, and outside of lab environments.
If the system will need to be calibrated occasionally, these readings
could help us understand how frequently such calibration sessions
might need to take place

\subsection{Evaluating a Closed-loop Passthought Authenticator}
Toward this end, propose to build a working, real-time passthought authenticator.
After performing an initial calibration setup, similar to our pilot study's protocol, above,
we will send subjects home with their custom-fit ear EEG, and ask them to use the device to authenticate
two to three times a day for a week.
(We may ask subjects to set a recurring event on their calendar to remind them, 
or pay subjects in accordances with how frequently they authenticated, up to some reasonable maximum).

After a week, subjects will return to the lab to return their equipment, debrief about their experience, and recieve payment. We will use this opportunity to walk subjects through a relatively detailed post-interview,
during which we will focus on their experiences with the device. 
The goal of this work is to understand both the technical issues facing ear-EEG authentication
(Does the system work reliably? When and why does it fail?), and the usability issues
(How do subjects feel about having an EEG device in their ear throughout the day?).
Finally, this closed-loop BCI system could help us understand how learning effects
on the human side might impact authentication performance, as the human and machine
co-adapt during multiple authentication attempts.

\subsection{Adding a Third Factor of Authentication}
After testing that our system can reliably provide authentication with a 
a secret passthought (knowledge) and unique EEG signal (inherence),
we will turn our attention to the final, third factor: posession
(of the unique, custom-fit ear EEG device).
We propose to cryptographically assure device posession
by assigning each custom-fit electrode a unique public/private keypair. 
User accounts could be associated with an earpiece's public key,
and authentication attempts from the earpiece could be signed its
its private key, producing authentication attempts that are not
forgeable by other custom-fit earpieces.

\section{Conclusion}
At the end of the proposed project,
we will have developed an authentication attempt that provides three factors
(knowledge, inherence and posession) in a single authentication attempt.
This work could lead to authentication strategies that are both more secure
and more useable than traditional, password-based authentication.

@@

[25] D. Garrett, D. Peterson, C. Anderson, and M. Thaut, “Comparison
of linear, nonlinear, and feature selection methods for eeg
signal classification,” IEEE Transactions on Neural Systems and
Rehabilitation Engineering, vol. 11, no. 2, pp. 141–144, Jun. 2003.
[Online]. Available: \url{http://ieeexplore.ieee.org/lpdocs/epic03/wrapper}.
htm?arnumber=1214704
\hspace{0pt} \\
\end{document}

>>>>

We propose the use of custom-fit Ear EEG technology as the platform for
investigating the feasibility, performance, and usability of one-step three-factor
authentication. In addition to the same knowledge factor and inherence factor as
in previous work, the user can submit in the same step the possession factor
in the form of the EEG-sensing ear-piece(s) that are custom-fitted to and worn in
their ear. These earpieces can serve as physical tokens in the same way as bank
ATM cards and wearable hardware tokens. Furthermore, because the earpieces
are custom-fitted to each individual, they will likely not be able to produce good
electrical impedances when worn by a different individual.

@@

We propose an update and expansion of the authentication tasks, to take
advantage of our lessons learned from previous studies on the relative strengths
of different mental tasks, with regards to authentication accuracy and usability as
reported by participants. Furthermore, given the in-ear placement of the
electrodes, and hence spatial proximity to the temporal lobes, we see a unique
opportunity to design and test novel authentication tasks based on either auditory
imagery or auditory steady-state response (ASSR). The ten authentication tasks and their
attributes are listed in tables 1 and 2 below.

@@

We will quantify the security and usability performances of these
tasks, both collectively as they contribute to overall authentication accuracy, as
well as comparatively against one another.
We will use OpenBCI, a open-source EEG biosensing system as our data
collection platform. OpenBCI, with a price tag of about \$600,
is an affordable alternative to medical-grade EEG systems that can
cost tens of thousands of dollars. Recent work has demonstrated the robustness 
of OpenBCI compared to an industry standard medical-grade EEG system, 
particularly for non-medical use cases such as the one we propose [26].
Using OpenBCI requires integrating the custom-fit Ear EEG devices with
the OpenBCI board. We will also need to extend our own data collector
software, to support the recording of OpenBCI data that is synchronized with
PsychoPy, which we use for stimuli presentation. We will recruit on the order of
20-30 participants who are willing to commit to multiple session visits, first for ear
molding, then for device impedance testing, and then for one or more data
collection sessions.

@@

Our initial participants were recruited from a nearby university and scheduled for ear molding
and impedance checking sessions. Finally, the data collection visit was scheduled and took
approximately 90 minutes for set up and experiment execution. The OpenBCI system we used
allows for 8 channels of simultaneous recording, along with separate ground and reference channels.
Data was initially collected with the ground placed at the center of the forehead, and using the left
mastoid as reference, though we can easily re-reference to another channel by subtracting a desired
channel (such as right mastoid). Each earpiece (shown in the image below) contain three channels: 
one placed on the helix, and two inside the canal - one front-facing and the other back-facing. The remaining
two channels were placed on the right mastoid for later re-referencing, and at approximately Fp1 (on the 
forehead above the left eye) for validating the data collected in the ears against a scalp-based measure. 
Before beginning the experiment, the data from all channels was visualized and participants were asked to
blink and clench their jaws to confirm visibly that all channels were active and properly connected.

During the experiment, participants were seated in a comfortable position in a quiet room facing a laptop screen on which the
instructions and stimuli were presented using PsychoPy. Each task was completed once in sets five trials each, and then
each was completed again for another five trials. Each trial was 10 seconds in length, for a total of 10 trials and 
100 seconds of data collected per task. The instructions were read aloud to the participant by the experimenter, and
the experiment was advanced using a pointer held in the participant's lap to minimize motion artifacts in the data.  
The experimenter also recorded the participant's chosen secrets for the sport, song, face, speech, and sequence
 tasks and reminded the participant of these for the second set of trials.

@@

The recorded impedances of the earpiece electrodes were less than 5 kOhms except one, a benchmark used widely in previous
ear EEG work. The left helix electrode of one participant was measured at 9 kOhms, and generally the 
helix impedances for both participants were higher than their ear canal counterparts. We expected this result, given that the
helix electrode relies on quality of the earpiece's fit outside the ear for good contact, and is not as securely and tightly placed as the
electrodes within the ear canal. Nonetheless, the data from all electrodes were tested in the remaining two data quality tests.

For the alpha-attenuation test, data from the "Breathe" task was compared with that of the "Breathe - Open" task. It is a well-
known feature of EEG data that activity in the alpha-band (approximately 8-12 Hz range) increases when the eyes are closed
compared with a similar state with eyes open. For both of our pilot participants this attenuation is clearly visible even in just
a single trial's data. To further validate, we also performed this calculation on the data collected from the Fp1 electrode and see
the effect clearly here as well. It is important to note that the left ear results are reported using the right mastoid as reference, and
the right ear results in turn using the left mastoid as reference. When using the same side mastoid for reference the effect is not 
visible, though it may be if we average across many trials. This is not surprising, as the further a reference electrode is from the active
channel the less "real" signal is being subtracted from the active channel. This has important design implications for eventual 
real-world deployment of this authentication method however, as it will likely require pieces worn on or around both ears to properly 
function, and not just one. The figures below show the alpha attenuation in the left and right ear channels, as well as Fp1.

@@

\begin{figure}[h]
  \centering
  \begin{minipage}[b]{0.45\textwidth}
    \includegraphics[width=\textwidth]{001_ASSR_Left.jpg}
  \end{minipage}
  \hfill
  \begin{minipage}[b]{0.45\textwidth}
    \includegraphics[width=\textwidth]{001_ASSR_Right.jpg}
  \end{minipage}
\caption{Power spectra for data collected from the earpiece channels during 40 Hz ASSR stimulus. Again, the spike is clearly visible though not at 40 Hz, however it 
does match the activity measured at Fp1.}
\end{figure}

@@

Following [5], we use logarithmic binning to produce
compressed feature vectors of a variable size. This technique
has been show to offer robust, linear classifiability in healthy
subjects. It is unique in its use of the entire frequency
spectrum. Since EEG activity is primarily associated with frequencies
from 1-40Hz, we presume this range contains the majority of
relevant signal. However, we do not rule out the possibility
that useful signal exists in other frequency ranges. Slight muscular
activity, for example, might be correlated with mental gestures
in some cases. Logarithmic binning produces feature vectors
biased toward known sources of signal, while still including
data points from outside this frequency range that may be
informative.

@@

Though our classifier's accuracy is strong, we still cannot say that it relies 
on both knowledge and inherence to authenticate the subject. Is it the passthought
that the classifier is authenticating, or do users (or earpieces) simply have characteristic
EEG readings?

@@

In expanding on the feasibility study we describe above, 
our first task comprises of collecting data from more subjects,
ideally for a subject population of 30.
In addition, we hope to collect more diverse data,
by taking regular samples over a period of a few weeks,
and in a variety of circumstances. Our goal here is to gauge the stability
of EEG readings over time, and outside of lab environments.
If the system will need to be calibrated occasionally, these readings
could help us understand how frequently such calibration sessions
might need to take place.

\subsection{Evaluating a Closed-loop Passthought Authenticator}
Toward this end, propose to build a working, real-time passthought authenticator.
After performing an initial calibration setup, similar to our pilot study's protocol, above,
we will send subjects home with their custom-fit ear EEG, and ask them to use the device to authenticate
two to three times a day for a week.
(We may ask subjects to set a recurring event on their calendar to remind them, 
or pay subjects in accordance with how frequently they authenticated, up to some reasonable maximum).

After a week, subjects will return to the lab to return their equipment, debrief about their experience, and receive payment. We will use this opportunity to walk subjects through a relatively detailed post-interview,
during which we will focus on their experiences with the device. 
The goal of this work is to understand both the technical issues facing ear-EEG authentication
(Does the system work reliably? When and why does it fail?), and the usability issues
(How do subjects feel about having an EEG device in their ear throughout the day?).
Finally, this closed-loop BCI system could help us understand how learning effects
on the human side might impact authentication performance, as the human and machine
co-adapt during multiple authentication attempts.

\subsection{Adding a Third Factor of Authentication}
After testing that our system can reliably provide authentication with a 
a secret passthought (knowledge) and unique EEG signal (inherence),
we will turn our attention to the final, third factor: possession
(of the unique, custom-fit ear EEG device).
We propose to cryptographically assure device possession
by assigning each custom-fit electrode a unique public/private keypair. 
User accounts could be associated with an earpiece's public key,
and authentication attempts from the earpiece could be signed its
its private key, producing authentication attempts that are not
forgeable by other custom-fit earpieces.

\section{Conclusion}
At the end of the proposed project,
we will have developed an authentication attempt that provides three factors
(knowledge, inherence and possession) in a single authentication attempt.
This work could lead to authentication strategies that are both more secure
and more useable than traditional, password-based authentication.

@@

[25] D. Garrett, D. Peterson, C. Anderson, and M. Thaut, “Comparison
of linear, nonlinear, and feature selection methods for eeg
signal classification,” IEEE Transactions on Neural Systems and
Rehabilitation Engineering, vol. 11, no. 2, pp. 141–144, Jun. 2003.
[Online]. Available: \url{http://ieeexplore.ieee.org/lpdocs/epic03/wrapper}.
htm?arnumber=1214704
\hspace{0pt} \\

[26] Frey, Jeremy. "Comparison of an open-hardware electroencephalography amplifier with medical
grade device in brain-computer interface applications." arXiv preprint arXiv:1606.02438 (2016).
\hspace{0pt} \\

\end{document}
